---
title: "Tuning POPMAPS parameters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tuning POPMAPS parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

This vignette shows the recommended `popmaps2` parameter-tuning workflow. The
goal is not to find universal defaults. The goal is to test whether a parameter
combination predicts withheld empirical ancestry estimates well for the species,
sampling design, and interpolation surface at hand.

Lower RMSE, MAE, and Hellinger distance are better. Higher dominant ancestry
accuracy and dominant-axis support are better.

## Setup

```{r setup}
library(popmaps2)

data(hija_raster)
data(hija_struc)

ex_raster <- raster::aggregate(hija_raster, fact = 240)
```

## Suggest A Starting Grid

`suggest_tuning_grid()` uses empirical sampling-site distances to suggest a
transparent starting grid for geographic-distance interpolation. The suggested
`popmod` values are calculated from interpretable retained weights at a reference
distance.

```{r suggest-grid}
grid <- suggest_tuning_grid(hija_struc)
grid
```

These values should be treated as a starting point. They are useful because they
scale to the empirical sampling design instead of relying on one fixed parameter
set for every species.

## Leave-One-Site-Out Tuning

Leave-one-site-out validation withholds each empirical site, predicts its
ancestry coefficients from the remaining sites, and scores the prediction.

```{r loo-tuning}
tuning <- tune_popmaps(
  input_raster = ex_raster,
  input_locs = hija_struc,
  surface = "G",
  empirical_pt_dist = grid$empirical_pt_dist,
  num_sites = grid$num_sites,
  num_tested = grid$num_tested,
  popmod = grid$popmod,
  quiet = TRUE
)

tuning$best
```

The `distance_units` column records whether distance summaries are geographic
kilometers (`surface = "G"`) or least-cost distance units (`surface = "C"`).
The unit-neutral columns `half_distance` and `ten_pct_distance` describe the
distance at which the distance-decay weight is 0.5 and 0.1.

## Diagnose Support

`diagnose_tuning()` helps distinguish strong support from a broad set of
near-best alternatives. This matters because a management surface should not
look more certain than the empirical data support.

```{r diagnose}
diagnostics <- diagnose_tuning(tuning)

diagnostics$overview
diagnostics$parameter_ranges
```

If many combinations are near-best, report that uncertainty instead of choosing
a single row visually.

## Spatial-Block Validation

Spatial-block validation is stricter than leave-one-site-out validation because
it asks whether parameters predict into spatially grouped withheld sites.

```{r spatial-block}
spatial_tuning <- tune_popmaps(
  input_raster = ex_raster,
  input_locs = hija_struc,
  surface = "G",
  validation = "spatial_block",
  n_blocks = 4,
  spatial_block_repeats = 3,
  spatial_block_seed = 1,
  empirical_pt_dist = grid$empirical_pt_dist,
  num_sites = grid$num_sites,
  num_tested = grid$num_tested,
  popmod = grid$popmod,
  quiet = TRUE
)

spatial_tuning$best
```

Repeated spatial blocks are especially useful when sampling locations are
clustered, because they expose whether the selected parameters depend strongly
on one spatial partition.

## Adaptive Exploration

`adaptive_tune_popmaps()` samples parameter space with random or Latin-hypercube
sampling, then optionally refines around the best-performing region. It is a
practical alternative when the full parameter grid would be too large.

```{r adaptive}
adaptive <- adaptive_tune_popmaps(
  input_raster = ex_raster,
  input_locs = hija_struc,
  n_initial = 8,
  n_refine = 4,
  seed = 1,
  quiet = TRUE
)

adaptive$best
```

Adaptive tuning should still be interpreted with diagnostics. A sampled search
can miss narrow optima, so use it as a way to explore parameter space before
committing to a more focused validation run.

## Mapping A Tuned Run

After choosing a parameter set, run `popmaps()` once with those values and
export both analysis rasters and map figures. The map products should show the
same uncertainty that drove parameter selection: broad low-confidence regions
are useful information, not a plotting failure.

```{r tuned-map}
best <- tuning$best[1, ]

aps <- popmaps(
  input_raster = ex_raster,
  input_locs = hija_struc,
  surface = "G",
  empirical_pt_dist = best$empirical_pt_dist,
  num_sites = best$num_sites,
  num_tested = best$num_tested,
  popmod = best$popmod,
  threshold = 0,
  ncore = 1
)

write_popmaps(
  pop_raster_list = aps,
  input_raster = ex_raster,
  dir = "tuned-popmaps-output",
  prefix = "hija-tuned",
  overwrite = TRUE
)

write_popmaps_plot(
  pop_raster_list = aps,
  input_raster = ex_raster,
  path = "tuned-popmaps-output/hija-tuned-map.png",
  input_locs = hija_struc,
  type = "confidence",
  style = "manuscript",
  background_threshold = 0.1015,
  overwrite = TRUE
)
```

Use `type = "boundary"` for hard ancestry assignments and `type = "axis"` with
`axis = 1`, `axis = 2`, and so on to inspect individual ancestry-axis surfaces.
Those views answer different questions and are often worth saving together.
