---
title: "Quick start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

This vignette walks through the smallest complete `popmaps2` workflow:

1. prepare empirical ancestry locations and a prediction raster;
2. tune a small parameter grid;
3. run POPMAPS with the selected parameters;
4. convert the output to modern raster objects;
5. write GeoTIFF layers and a map figure.

The example uses the small bundled *Hesperidanthus jaegeri* data so it can run
quickly after installation. It is a learning workflow, not a recommended
analysis scale for a publication or management decision.

## Run The Bundled Script

After installing `popmaps2`, the fastest check is to source the bundled
start-here script:

```{r run-script}
library(popmaps2)

start_here <- system.file("examples", "start-here.R", package = "popmaps2")
source(start_here)
```

The script writes its outputs to a temporary directory and prints that path at
the end. For your own project, copy the workflow shape and change the output
directory to a project folder.

## Load Example Data

The package includes one small empirical example:

```{r load-data}
library(popmaps2)

data(hija_raster)
data(hija_struc)

ex_raster <- raster::aggregate(hija_raster, fact = 240)
input_locs <- hija_struc
```

`hija_raster` supplies the prediction grid. For `surface = "G"`, raster values
are ignored for distance calculations; the raster supplies geometry and any
`NA` prediction mask. For `surface = "C"`, raster values define conductance or
cost distance across the landscape.

`hija_struc` is an empirical ancestry table. The required column order is:

| Column | Meaning |
| --- | --- |
| 1 | Site name |
| 2 | Longitude or x-coordinate |
| 3 | Latitude or y-coordinate |
| 4...n | Ancestry coefficients |

Before modeling starts, `popmaps2` checks for common input problems: blank or
duplicated site names, non-numeric coordinates, invalid ancestry values, points
outside the raster extent, and points falling on `NA` raster cells.

## Tune Parameters

Use `tune_popmaps()` to evaluate candidate parameters against withheld empirical
ancestry estimates. Lower RMSE, MAE, and Hellinger distance are better. Higher
dominant ancestry accuracy and dominant ancestry probability are better.

```{r tune}
tuning <- tune_popmaps(
  input_raster = ex_raster,
  input_locs = input_locs,
  surface = "G",
  empirical_pt_dist = c(0, 5),
  num_sites = c(5, 8),
  num_tested = c(2, 3),
  popmod = c(-0.01, -0.05),
  threshold = 0,
  quiet = FALSE
)

tuning$best
```

This grid is intentionally tiny. For real analyses, use a biologically
motivated grid, `suggest_tuning_grid()`, `adaptive_tune_popmaps()`, or the
surface-specific tools described in the tuning and surface-comparison
vignettes.

## Run POPMAPS

The best tuning row provides a transparent starting point for a final map run:

```{r run-popmaps}
best <- tuning$best[1, ]

aps <- popmaps(
  input_raster = ex_raster,
  input_locs = input_locs,
  surface = "G",
  empirical_pt_dist = best$empirical_pt_dist,
  num_sites = best$num_sites,
  num_tested = best$num_tested,
  popmod = best$popmod,
  threshold = 0,
  quiet = FALSE
)
```

Use `quiet = FALSE` when you want progress messages during longer runs.

## Export Rasters

`popmaps()` preserves the original POPMAPS list output. Use `popmaps_rast()` for
modern `terra` workflows and `write_popmaps()` for GeoTIFF output:

```{r export-rasters}
aps_raster <- popmaps_rast(aps, ex_raster)
aps_raster

output_dir <- file.path(tempdir(), "popmaps2-quick-start")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

written_layers <- write_popmaps(
  pop_raster_list = aps,
  input_raster = ex_raster,
  dir = output_dir,
  prefix = "hija",
  overwrite = TRUE
)

written_layers
```

The output layers are:

| Layer | Meaning |
| --- | --- |
| `boundary` | Hard ancestry assignment |
| `ancestry_probability` | Confidence in the hard assignment |
| `axis_*` | Weighted ancestry estimate for each ancestry axis |

## Export A Map Figure

Use `write_popmaps_plot()` for manuscript-style figures:

```{r export-map}
map_path <- file.path(output_dir, "hija-ancestry-map.png")

write_popmaps_plot(
  pop_raster_list = aps,
  input_raster = ex_raster,
  path = map_path,
  input_locs = input_locs,
  style = "manuscript",
  type = "ancestry",
  overwrite = TRUE
)

map_path
```

Use `type = "boundary"` for hard assignments and `type = "axis"` with `axis = 1`,
`axis = 2`, and so on for individual ancestry-axis surfaces.

## What To Read Next

Read the other vignettes when the basic workflow runs cleanly:

- `vignette("parameter-tuning", package = "popmaps2")` explains tuning,
  diagnostics, spatial-block validation, and adaptive search.
- `vignette("surface-comparison", package = "popmaps2")` explains geographic,
  suitability, conductance, resistance, EEMS, and FEEMS surface workflows.
- `vignette("local-empirical-validation", package = "popmaps2")` explains local
  validation scripts for larger empirical examples that should not live inside
  the package repository.
