---
title: "Comparing POPMAPS interpolation surfaces"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing POPMAPS interpolation surfaces}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette shows the intended surface-comparison workflow in `popmaps2`.
The goal is not to fit species distribution models, EEMS, FEEMS, or resistance
models inside `popmaps2`. The goal is to take candidate surfaces from those
outside analyses, use them for ancestry interpolation, and ask which surface
best predicts withheld empirical ancestry estimates.

## Terminology

`surface = "G"` means geographic interpolation. Raster values are ignored for
distance calculations. The raster only defines the prediction grid, extent, cell
size, and optional missing cells.

`surface = "C"` means conductance or cost-distance interpolation. Raster values
matter because they define how easily ancestry information moves across the
landscape. Higher suitability or conductance values imply easier movement. If a
user supplies resistance values, `popmaps2` converts them to conductance with an
inverse transform.

`surface_from_points()` is mostly a convenience for `G`: it creates a simple
grid around empirical sampling coordinates when the user does not already have a
raster. It can create a constant `C` surface, but that is a neutral template,
not a biological landscape hypothesis.

## Starting From Empirical Locations

POPMAPS location tables have site name, longitude, latitude, and ancestry
coefficients:

```{r}
library(popmaps2)

data(hija_struc)
head(hija_struc)
```

If locations are stored as `sf` point features, convert them first:

```{r, purl = FALSE}
input_locs <- locs_from_sf(
  sf_points,
  site_col = "site",
  ancestry_cols = c("axis1", "axis2", "axis3")
)
```

If no raster is available, create a geographic prediction grid from the
coordinates:

```{r}
geographic_surface <- surface_from_points(
  points = hija_struc,
  resolution = 0.01,
  surface = "G"
)
```

For a real analysis, choose `resolution` in the coordinate units of the data.
If coordinates are longitude and latitude, that means decimal degrees. If the
data are projected, it may mean meters.

## Preparing Candidate Surfaces

An SDM logistic suitability raster can be treated as conductance:

```{r, purl = FALSE}
sdm_surface <- prepare_popmaps_surface(
  input_raster = sdm_raster,
  surface = "C",
  surface_values = "suitability"
)
```

A resistance raster can be supplied directly and inverted internally:

```{r, purl = FALSE}
resistance_surface <- prepare_popmaps_surface(
  input_raster = resistance_raster,
  surface = "C",
  surface_values = "resistance"
)
```

If a multi-layer raster contains several candidate surfaces, convert every layer
into a named list:

```{r, purl = FALSE}
candidate_surfaces <- surfaces_from_raster_stack(
  input_raster = multi_layer_raster,
  surface = "C",
  surface_values = c("suitability", "conductance", "resistance"),
  include_geographic = TRUE
)
```

EEMS and FEEMS outputs should be generated outside `popmaps2`, then imported as
raster-like outputs or coordinate/value tables:

```{r, purl = FALSE}
eems_surface <- surface_from_eems(
  input = eems_table,
  value_col = "migration",
  surface_values = "conductance"
)

feems_surface <- surface_from_feems(
  input = feems_table,
  value_col = "w",
  surface_values = "conductance"
)
```

## Comparing Surfaces

Surface comparison uses the same validation design across each candidate
surface. With `surface_grid = "surface_specific"`, missing tuning values are
suggested separately for each surface so geographic kilometers and least-cost
distance units are not forced onto the same scale.

```{r, purl = FALSE}
surfaces <- list(
  geographic = geographic_surface,
  sdm_suitability = sdm_surface,
  resistance = resistance_surface
)

comparison <- compare_popmaps_surfaces(
  input_locs = input_locs,
  surfaces = surfaces,
  validation = "spatial_block",
  n_blocks = 4,
  spatial_block_repeats = 5,
  spatial_block_seed = 1
)

comparison$summary
comparison$support
```

Lower RMSE, MAE, and Hellinger distance are better. Higher dominant ancestry
accuracy and dominant-axis support are better. The default support rule
labels a surface as clearly supported only when no other surface is within the
near-best tolerance.

## Reporting

Use the report helper to create CSVs, plots, and a short Markdown report:

```{r, purl = FALSE}
report <- write_surface_comparison_report(
  comparison,
  dir = "surface-comparison-report",
  prefix = "hija-surfaces"
)

report$report
report$figures
```

The report is meant to support interpretation, not automate judgment. A surface
that is "indistinguishable" from another candidate should usually be treated as
equally supported by the current data and validation design.

## Mapping The Selected Surface

Once a surface and parameter set are selected, use the same surface in
`popmaps()` and export the resulting rasters and map figures. This keeps the
validation decision connected to the management product.

```{r selected-surface-map, purl = FALSE}
best_surface <- comparison$best$surface_name[1]
selected_surface <- surfaces[[best_surface]]
selected_params <- comparison$best[1, ]

popmaps_args <- list(
  input_raster = selected_surface$raster,
  input_locs = input_locs,
  surface = selected_surface$surface,
  empirical_pt_dist = selected_params$empirical_pt_dist,
  num_sites = selected_params$num_sites,
  num_tested = selected_params$num_tested,
  popmod = selected_params$popmod,
  threshold = 0,
  ncore = 1
)

if (identical(selected_surface$surface, "C")) {
  popmaps_args$surface_values <- selected_surface$surface_values
}

aps <- do.call(popmaps, popmaps_args)

write_popmaps_plot(
  pop_raster_list = aps,
  input_raster = selected_surface$raster,
  path = "selected-surface-map.png",
  input_locs = input_locs,
  type = "confidence",
  style = "manuscript",
  overwrite = TRUE
)
```

For suitability or conductance surfaces, a `background_threshold` can be added
to show confidence or ancestry-axis values only in cells treated as occupied or suitable.
That is a visualization mask, not a new model fit.

## Practical Guidance

Start with a small candidate set:

- a geographic `G` surface;
- one SDM or habitat suitability `C` surface;
- one resistance or conductance surface if the species has a defensible
  landscape hypothesis;
- EEMS or FEEMS exports only when those analyses have already been run and
  interpreted upstream.

Prefer spatial-block validation when the management question involves
prediction into unsampled areas. Leave-one-out validation is useful for checking
local interpolation behavior, but it can be optimistic when nearby empirical
sites are very similar.
